---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


***


```{r results='hide'}
##library(dplyr)

```


##Loading and preprocessing the data 
***

```{r}
act <- read.csv("activity.csv")
options(scipen = 1, digits = 2)
```


##What is mean total number of steps taken per day?
***

```{r}

per_day <- tapply(act$steps, act$date, sum)

hist(per_day, main = "Histogram of Total Steps per Day", xlab = "Steps per Day")

mn <- round(mean(per_day, na.rm = T), digits = 2)

md <- median(per_day, na.rm = T)

```




####The MEAN total number of steps per day is `r mn`  


####The MEDIAN total number of steps per day is `r md`



##What is the average daily activity pattern?
***

```{r}
#use ggplot stat_summary to plot mean steps across all days
library(ggplot2)
require(ggplot2)
#act$ts <- formatC(act$interval, digits = 3, flag = 0)
#act$ts <- gsub("^(\\d{2})(\\d{2})$", "\\1:\\2", act$ts)
g <- ggplot(data = act, aes(act$interval, act$steps))
g <- g + stat_summary(fun.y = mean, geom = "line", na.rm = T)
g <- g + xlab("Interval") + ylab("Mean Steps")
g

#get the max activity interval
act_ts <- reshape(act, idvar = "date", timevar = "interval", direction = "wide")
mns <- colMeans(act_ts[,-1], na.rm = T)
max_val <- max(mns)


#get the column name and format it as a time 
max_col <- names(which.max(mns))
max_col <- substr(max_col,7,10)
max_col <- formatC(as.numeric(max_col), digits = 3, flag = 0)
max_col <- gsub("^(\\d{2})(\\d{2})$", "\\1:\\2", max_col)


```


####The maximum activity occurs in the `r max_col` interval when the average steps taken are `r max_val`.

##Imputing missing values
***

####Strategy
The strategy used to impute NA values is to take the mean steps for the particular interval in question across all the days. So, for example, for the 8:00-8:05 interval, the mean for this interval across all days was used to fill in any NA values seeen for this interval.

``` {r}
# get the total na values
tot_na <- table(is.na(act$steps))[2]

# impute the median values and replace NA values
library(data.table)
act_dt <- data.table(act)
act_dt[,steps := ifelse(is.na(steps), as.integer(round(mean(steps, na.rm = T))), steps), by = interval]

per_day <- tapply(act_dt$steps, act_dt$date, sum)

hist(per_day, main = "Histogram of Total Steps per Day", xlab = "Steps per Day")

mn2 <- round(mean(per_day, na.rm = T), digits = 2)

md2 <- median(per_day, na.rm = T)



```

There are `r tot_na` missing observations in the dataset. The MEAN total number of steps per day with imputed missing values is `r mn2`. The difference between the non-imputed NA values and the imputed NA values is only `r mn - mn2` steps. The MEDIAN total number of steps per day is `r md2` steps. The difference in median values is only `r md - md2` steps. The effect of imputing the values is negligible as expected because the percent of NA values was fairly low to begin with at `r tot_na/nrow(act) * 100`%.



##Are there differences in activity patterns between weekdays and weekends?
***
```{r}

act_dt$wkdy <- "Weekday"
act_dt[,wkdy := ifelse(weekdays(as.Date(date)) %in% c("Saturday", "Sunday"), "Weekend", "Weekday")]
ft <- factor(act_dt$wkdy)
g <- ggplot(data = act_dt, aes(act$interval, act$steps))
g <- g + stat_summary(fun.y = mean, geom = "line", na.rm = T)
g <- g + xlab("Interval") + ylab("Mean Steps")
g <- g + facet_grid(wkdy~.)
g

```

Yes, it appears that the spike in activity between 8-10 on weekdays is not there on weekends. Also the weekend activity is more constant and at ahigher level throughout the day.

